---
title: "CHR"
format: html
editor: visual
---

## 🌐 🌽 County Health Rankings w/ Bayesian Regression

### Libraries

```{r}
library(tidyverse)
library(janitor)
library(rstanarm)
library(here)
library(readxl)
```

### Data

```{r}
chr_data <- read_excel(here("data/2025_County_Health_Rankings_Data.xlsx"), sheet = 2, skip =1)
```

```{r}
chr_data <- chr_data %>%
  clean_names()
```

#### Regression Variables of Interest

-   `food_environment_index`

    -   The **Food Environment Index (FEI)** is a **County Health Rankings** composite (0–10 scale, where **10 = best**) that summarizes how easy it is for people in a county to get enough **healthy, affordable food**.

        It combines two pieces of information:

        1.  **Food insecurity** – the estimated share of residents who lack reliable access to enough food.

        2.  **Limited access to healthy foods** – the share of people who are **low-income and far from a grocery store** (typically ≥1 mile in urban areas or ≥10 miles in rural areas).

            These components are blended and rescaled to 0–10:

            **Higher FEI (closer to 10)** → lower food insecurity and better physical access to healthy food.

            **Lower FEI (closer to 0)** → more food insecurity and poorer access (more “food deserts”).

-   `percent_fair_or_poor_health`

    -   a County Health Rankings (CHR) indicator from the BRFSS survey:

        -   **What it is:** the **percentage of adults** who answer **“fair” or “poor”** to the question, “In general, would you say your health is excellent, very good, good, fair, or poor?”

Simplify dataframe to select columns of interest

```{r}
chr_data_simplified <- chr_data %>%
  select(state, county,
         food_environment_index, percent_fair_or_poor_health)
summary(chr_data_simplified)
head(chr_data_simplified)

```

### Simple scatter plot, linear model

```{r}
ggplot(chr_data_simplified, aes(x= food_environment_index, y = percent_fair_or_poor_health)) +
  geom_point(alpha = 0.1) +
  geom_smooth()
```

```{r}
lm(percent_fair_or_poor_health ~ 
     food_environment_index, chr_data_simplified)
```

### Priors

Literature-Informed Priors: Fair/Poor Health \~ Food Environment Index

Empirical Background

Several studies using BRFSS, County Health Rankings (CHR), or linked datasets find that better local food access is associated with lower rates of poor self-rated health.

| Parameter | Prior (percent scale) | What it encodes | Literature tie-in |
|----|----|----|----|
| **Intercept (β₀)** | `Normal(17, 4)` | At **average FEI**, a typical county has \~**17%** fair/poor health (95% prior ≈ **9–25%**) | BRFSS/CHR descriptive range (≈14–20% typical); compatible with Berkowitz (food insecurity ↔ higher % fair/poor) |
| **Slope (β₁)** | `Normal(-1.3, 0.6)` *(pp per +1 FEI)* | Each **+1 FEI** is expected to **reduce** fair/poor health by **\~1.3 percentage points** on average (95% prior ≈ **−2.5 to −0.1** pp) | **Fan & Jin (2022)** report −**1.3 pp** per +1 FEI; **Liese et al. (2014)** (odds ≈1.1× worse per unit) implies ≈ **−1 to −1.6 pp** per +1 FEI near 15–20% baseline |
| **Residual SD (σ)** | `Exponential(1/4.5)` | Mean residual scatter ≈ **4.5 percentage points** (allows substantial county heterogeneity) | County-level BRFSS/CHR models often show **4–6 pp** residual SD after single-predictor adjustment |

### Simulation with rstanarm

```{r}
health_model <- stan_glm(percent_fair_or_poor_health ~ food_environment_index, data = chr_data_simplified,
                       family = gaussian,
                       prior_intercept = normal(17, 4),
                       prior = normal(-1.3, 0.6), 
                       prior_aux = exponential(1/4.5),
                       chains = 4, iter = 5000*2, seed = 84735)
```

```{r}
glimpse(chr_data_simplified)
```

→ Each **+1** in **Food Environment Index** is associated with **\~2.6 percentage-point lower** fair/poor health, on average.

**Residual SD (sigma)** = **3.6 pp**: after accounting for FEI, counties typically vary by about ±3–4 percentage points around the line.

### Challenge

Make a map describing the **Food Environment Index** in the US

```{r}
library(ggplot2)
library(dplyr)
library(stringr)
library(maps)
library(viridis)
library(usmap)
```

```{r}
df <- chr_data_simplified   # your data (state, county, food_environment_index, percent_fair_or_poor_health)

# Build the 'polyname' key maps uses: "state,county" all lowercase
dat_fips <- df %>%
  filter(!is.na(county)) %>%                         # drop statewide rows
  mutate(
    polyname = paste(str_to_lower(state), str_to_lower(county), sep = ","),
    # normalize 'St.' to 'saint ' to match maps naming
    polyname = str_replace(polyname, "^st\\.?\\s+", "saint ")
  ) %>%
  left_join(maps::county.fips, by = "polyname") %>%  # adds 'fips' (integer)
  mutate(fips = sprintf("%05d", fips))               # zero-pad to 5 chars

```

```{r}
plot_usmap(data = dat_fips, regions = "counties", values = "food_environment_index") +
  scale_fill_viridis_c(name = "Food env.\nindex", na.value = "grey90") +
  labs(title = "Food Environment Index by U.S. County") +
  theme(legend.position = "right")

```

```{r}

chr_data <- chr_data_simplified

# Do we even have Alaska/Hawaii rows?
chr_data %>% filter(state %in% c("Alaska","Hawaii")) %>% 
  count(state, county, sort = TRUE) %>% 
  print(n = 50)

# Are the values missing for those rows?
df %>% filter(state %in% c("Alaska","Hawaii")) %>% 
  summarize(
    n_rows = n(),
    n_county_na = sum(is.na(county)),
    any_value_missing = any(is.na(food_environment_index) | is.na(percent_fair_or_poor_health))
  )

```

```{r}
library(tigris)
# light normalizer: remove suffixes like "County", "Borough", "Census Area", etc.
clean_county <- function(x){
  x %>% 
    str_to_lower() %>%
    str_replace_all("^st\\.?\\s+", "saint ") %>%
    str_replace_all("\\s+(county|parish|city and borough|borough|census area|municipality|city)$","") %>%
    str_squish()
}

# Build a fresh FIPS lookup from tigris
fips_xwalk <- tigris::fips_codes %>%
  transmute(
    state = str_to_lower(state_name),
    county = clean_county(county),
    fips   = paste0(state_code, county_code)
  ) %>%
  distinct(state, county, .keep_all = TRUE)

# Prepare your data and join to FIPS
dat_fips <- chr_data_simplified %>%
  filter(!is.na(county)) %>%               # drop statewide aggregates for county map
  mutate(
    state  = str_to_lower(state),
    county = clean_county(county)
  ) %>%
  left_join(fips_xwalk, by = c("state","county"))

# What failed to match?
missing_ak_hi <- dat_fips %>%
  filter(state %in% c("alaska","hawaii"), is.na(fips)) %>%
  distinct(state, county)
missing_ak_hi

```

```{r}
# light normalizer: remove suffixes like "County", "Borough", "Census Area", etc.
clean_county <- function(x){
  x %>% 
    str_to_lower() %>%
    str_replace_all("^st\\.?\\s+", "saint ") %>%
    str_replace_all("\\s+(county|parish|city and borough|borough|census area|municipality|city)$","") %>%
    str_squish()
}

# Build a fresh FIPS lookup from tigris
fips_xwalk <- tigris::fips_codes %>%
  transmute(
    state = str_to_lower(state_name),
    county = clean_county(county),
    fips   = paste0(state_code, county_code)
  ) %>%
  distinct(state, county, .keep_all = TRUE)

# Prepare your data and join to FIPS
dat_fips <- chr_data_simplified %>%
  filter(!is.na(county)) %>%               # drop statewide aggregates for county map
  mutate(
    state  = str_to_lower(state),
    county = clean_county(county)
  ) %>%
  left_join(fips_xwalk, by = c("state","county"))

# What failed to match?
missing_ak_hi <- dat_fips %>%
  filter(state %in% c("alaska","hawaii"), is.na(fips)) %>%
  distinct(state, county)
missing_ak_hi

```

```{r}
dat_fips <- chr_data_simplified %>%
  filter(!is.na(county)) %>%
  mutate(
    state  = str_to_lower(state),
    county = str_to_lower(county),
    county = case_when(
      state == "alaska" & county %in% c("kusilvak", "kusilvak census area") ~ "wade hampton",
      state == "alaska" & county == "petersburg borough" ~ "petersburg",
      state == "alaska" & county == "wrangell city and borough" ~ "wrangell",
      TRUE ~ county
    ),
    county = clean_county(county)
  ) %>%
  left_join(fips_xwalk, by = c("state","county"))

```

```{r}
# Food Environment Index
plot_usmap(data = dat_fips, regions = "counties", values = "food_environment_index") +
  scale_fill_viridis_c(name = "Food env.\nindex", na.value = "grey90") +
  labs(title = "Food Environment Index by U.S. County") +
  theme(legend.position = "right")

# % Fair/Poor Health
plot_usmap(data = dat_fips, regions = "counties", values = "percent_fair_or_poor_health") +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "% fair/poor", na.value = "grey90") +
  labs(title = "Self-Reported Fair/Poor Health by U.S. County") +
  theme(legend.position = "right")

```
